A method for defining dispersed community territories

### 2024-03-02 manuscript/ras-reviews-response.md

Reviewers provided important and insightful comments that helped us improve this
paper considerably. Perhaps the biggest revision is that we now include some
limited sensitivity analysis to evaluate what happens when parameter values are
changed. As two of the reviewers raised concerns about the resolution of the
grid, tests were also conducted to see if higher resolutions would have a
noticeable impact on the results - they did not. Finally, two of the reviewers
asked that we add some broader implications to the discussion. This was done by
highlighting differences between the types of communities we are trying to
understand and other types of communities (religious, linguistic, etc.) that
archaeologists might find relevant to their research.

Note: minor formatting and spelling errors identified by the reviewers were
corrected in the text but are not discussed here.

### Response to Reviewer 1

1.  Distance traveled should be measured with hypotenuse.

    Response: The reviewer is correct about this, so the analysis was updated to
    account for it. It required only minor changes to the R code. We also note
    that the effect was minimal as a maximum slope was defined to constrain
    travel paths, so the difference in travel time ended up being less than a
    minute in most cases.

2.  Use 5th percentile in Lorentz curve as recommended by Campbell.

    Response: This was a major oversight on our part that we have now corrected,
    and we thank the reviewer for bringing it to our attention. As requested, we
    also clarify that we are using the 5th percentile of the Lorentz curve in
    text.

3.  Aggregation of elevation model grid cells biases estimates.

    Response: The concern is valid, so we conduct an additional sensitivity
    analysis at four different levels of aggregation (from no aggregation to the
    level of aggregation used in the paper). Descriptions of this were added to
    the methods and results. To summarize, we found that increasing resolution
    beyond that used in our paper did not have a meaningful impact on the
    results.

4.  Bi-directional Dijkstra forces isotropy.

    Response: In bi-directional search on a directed graph, search from the
    destination node uses edge weights in the direction of the destination node,
    not edge weights in the direction of the origin node as the reviewer seems
    to be implying, so anisotropy is preserved. At any rate, it appears to be a
    moot point since `{cppRouting}` uses regular old Dijkstra's algorithm when
    `get_distance_matrix()` is applied to an uncontracted graph (the method used
    here). So, what the reviewer wants is what we actually did. We just
    described it wrong. This has been corrected in the text.

5.  Reviewer proposes alternative clustering strategy.

    Response: The proposed alternative strategy for defining communities is an
    intriguing one, but not fundamentally different than what we propose. There
    are three main differences.

    \(i\) Rather than searching the entire grid, search in our algorithm ends
    when all sites are reached, which reduces the time complexity.

    \(ii\) The proposed alternative doesn't specify a rule for joining
    communities, which we consider to be one of the two key contributions of our
    algorithm (the other is weighting distance by population size).

    \(iii\) If i understand it correctly, the proposed alternative for defining
    boundaries is to use the smallest isochrone around all the centers in a
    community that encompasses all the nearby farm sites. This is an intriguing
    suggestion that deserves further consideration, and we would love to see a
    follow-up paper that explores it more. We note, however, that it would not
    guarantee concavity in the community polygon with respect to time - meaning
    the shortest path between two sites in the community would not be guaranteed
    to remain in the community. Of course, it could be argued that that's a
    difference that doesn't actually make a difference, that these are just
    crude estimates anyway. Our argument would be that interactions did not
    occur only at community centers, so the shortest path between all sites is
    probably relevant to defining the spatial extent of these farming
    communities. If the people living on these farms were routinely visiting
    their neighbors, the routes they took would have formed an important element
    of their spatial or geographic community.

### Response to Reviewer 2

1.  Include additional citations in methods.

    Response: Done! See especially the section on defining community centers.

2.  Discuss broader connections outside the regional focus.

    Response: We add two paragraphs to the discussion elaborating on the type of
    community we are delineating and how that differs from other types of
    communities that may be of interest to archaeologists and anthropologists,
    focusing in particular on different dimensions of community organization -
    biological relatedness, shared metaphors, religion, language, etc.

3.  Add map with place names mentioned in text.

    Response: Done! This actually provides a lot of context that will hopefully
    help readers better interpret the results. In fact, any reference to
    specific locations in the text were updated to ensure that they used the
    labeled locations in the map. It was also an enjoyable exercise! Thanks for
    the request!

4.  Include some discussion of how visual landmarks might shape these
    communities.

    Response: This is an intriguing idea. Bernardini and Peeples (2017, "Sight
    communities," *Am Ant V*80 N2) provide a good example of this as they use
    visual landmarks and viewsheds to define "sight communities." That said, our
    goal here is to define the spatial extent of dispersed *farming*
    communities, not sight communities, so we focus on farmsteads that serve as
    indicators of the locations of farms and bind them together into communities
    using the centers at which farmers would have interacted on a fairly regular
    basis.

5.  Aggregation of elevation model grid cells biases estimates.

    Response: See response to reviewer 1.

6.  Clarify whether slope is degree slope or percent slope.

    Response: Campbell's hiking function is specified to work with degree slope
    by default. This was clarified in the text.

### Response to Reviewer 3

1.  Discuss broader connections outside the regional focus.

    Response: See response to Reviewer 2.

2.  Sensitivity analysis to evaluate model parameters.

    Response: Done! In general, we found that the algorithm behaved as expected,
    though we note some volatility in the Cibola data. This is probably owing to
    the geographic sparsity of that sample, particularly with respect to the
    ratio of farms to centers.

3.  Region-specific definitions of community centers.

    Response: To clarify our reasons for this, we added the following to the
    section on community centers:

    "For given differences in population density and settlement patterning, the
    frequency and intensity of interaction between individuals at centers is
    likely to vary by region, so it is reasonable to expect that community
    centers themselves (at least in so far as they play the role of centers)
    should vary in their size and composition across regions, too."

### 2024-03-02 manuscript/manuscript.qmd
multiple changes based on rev reqs

---
title: A method for defining dispersed community territories
date: today
author:
  - name: Kenneth B. Vernon
    email: Kenneth.Vernon@colorado.edu
    affiliations: 
      - id: colorado
        address: | 
          Center for Collaborative Synthesis in Archaeology,  
          Institute of Behavioral Science,  
          University of Colorado, Boulder,  
          1440 15th Street,  
          Boulder, CO 80302
    attributes:
        corresponding: true 
  - name: Scott G. Ortman
    affiliations:
      - ref: colorado
abstract: |
  The transition from dispersed to aggregated forms of settlement reflects a critical shift in the relative value of social and primary (food) modes of production. Investigating trade-offs between them, however, requires estimates of the extent of community territories, including their nearby arable land. Here we demonstrate a simple algorithm to do that. Our algorithm is analogous to that used to define core-based statistical areas for the US census, though instead of modern metropolitan areas, we rely on community centers (or areas of known and persistent interaction between unrelated individuals). We provide examples of our algorithm by applying it to archaeological sites in the central Mesa Verde, northern Rio Grande regions, and Cibola regions. Together, these data provide a comprehensive, high resolution record of prehistoric maize farming in the North American Southwest, representing varying levels of aggregation and dispersal through time.
keywords:
  - Aggregation
  - Core-Based Statistical Area
  - Community Detection
  - Intensification
target-journal: Journal of Archaeological Science
bibliography: bibliography.bib
linenumbers: false
watermark: false
format: 
  simple-pdf: default
execute: 
  echo: false
  error: false
  # warning: false
engine: knitr
---

@@

From a complex systems perspective, the main way these two settlement morphologies vary is with respect to which costs are being minimized. In the dispersed pattern, costs associated with primary (food) production by the household are being minimized, whereas in the aggregated pattern, it is costs associated with various forms of social production, from government and ritual to economic exchange and warfare. Presumably, choices regarding which sorts of costs to minimize are driven by the relative productivities of primary and social production, which is to say, their relative importance, broadly construed, for human well-being in a given context.

@@

The datasets we apply our algorithm to come from three regions in the US Southwest: Central Mesa Verde (CMV), the Northern Rio Grande (NRG), and the Cibola or Zuni region (CIB) (see @fig-overview). The CMV and NRG datasets derive from Phase II of the Village Ecodynamics Project [@ortman2016; @schwindt_etal2016]. The CIB data derive from data syntheses produced by Schachner and Peeples [@peeples2018; @peeples_schachner2012; @schachner2012]. All of these data are now included in cyberSW [@mills_etal2020]. For summaries of the archaeological and environmental context in each region, see @tbl-regions.

![
Overview map showing the locations of farmsteads (black dots) and community centers (red triangles) in Central Mesa Verde, Northern Rio Grande, and Cibola.
](../figures/overview-maps.png){#fig-overview layout-align="center" width="100%"}

The CMV area encompasses the southwestern corner of Colorado, from the Mesa Verde cuesta in the south and east to the Dolores River in the north, and the Utah state line in the west. The elevation ranges from roughly 1,399.8 m to 3,032.8 m, with an average elevation of 1,999.0 m ($\sigma$ = 276.5 m). SKOPE [@bocinsky_etal2022] estimates suggest annual precipitation in the CMV study area has averaged around 39.5 cm ($\sigma$ = 9.1 cm) over the last 2,000 years, and it puts summer growing degree days for maize around 2,390.1 °F ($\sigma$ = 97.9 °F). Across most areas in the region, it is believed that direct precipitation was sufficiently high to support dry farming in most years [@bocinsky_kohler2014; @bocinsky_etal2016].

The NRG study area in north central New Mexico includes basically all of the Rio Grande valley between the Jemez Mountains to the west and the southernmost mountains of the Sangre De Cristo range to the east, with the Rio Grande itself running along the western edge of the valley at the foot of the Pajarito Plateau. The elevation ranges from roughly 1,587.9 m to 3,836.2 m, with an average elevation of 2,220.0 m ($\sigma$ = 371.3 m). SKOPE estimates the average annual precipitation in the NRG over the last 2,000 years to be around 40.6 cm ($\sigma$ = 8.2 cm), and it puts summer growing degree days for maize around 2,154.4 °F ($\sigma$ = 102.1 °F). It is argued that direct rainfall on the Pajarito Plateau made dry farming possible there, but in the valley farming required floodwater irrigation along watercourses [@duwe_anschuetz2013; @bocinsky_kohler2014; @bocinsky_etal2016; @ortman2016a].

The CIB region straddles the southern slope of the Colorado Plateau along the border between Arizona and New Mexico. Although this is a rugged terrain by any reasonable standard, it is notably less so than the other study areas. The elevation ranges from roughly 1,771.3 m to 2,777.1 m, with an average elevation of 2,145.9 m ($\sigma$ = 160.0 m). SKOPE estimates suggest annual precipitation in the CIB area over the last 2,000 years has averaged around 34.3 cm ($\sigma$ = 7.6 cm), and it puts summer growing degree days for maize around 2,317.3 °F ($\sigma$ = 93.8 °F). However, no area in the region receives sufficient direct rainfall to support dry farming, so some form of water management was required [@kintigh1985; @muenchrath_etal2002].

::: {#tbl-regions}

```{=latex}

\begin{singlespace}
\small
\begin{tabular}{lllll}
    & \textbf{Area [km2]} & \textbf{Dates [CE]} & \textbf{Archaeology} & \textbf{Environment} \\ 
    \hline  
    \noalign{\vskip2mm}  
  CMV & 4,556.4             & 725-1280            & \begin{tabular}[c]{@{}r@{}}Rooms: \hfill 67,746\\ Farms: \hfill 5,415\\ Centers: \hfill 512\end{tabular} & \begin{tabular}[r]{@{}r@{}}Precipitation: \hfill 39.5 cm ($\sigma$: 9.1)\\ Maize GDD: \hfill 2,390.1 $^{\circ}$F ($\sigma$: 97.9)\\ Elevation: \hfill 1,999.0 m ($\sigma$: 276.5)\end{tabular} \\[5mm]  
  \hline  
  \noalign{\vskip2mm}  
  NRG & 6,958.0             & 900-1550            & \begin{tabular}[c]{@{}r@{}}Rooms: \hfill 46,226\\ Farms: \hfill 2,316\\ Centers: \hfill 181\end{tabular} & \begin{tabular}[r]{@{}r@{}}Precipitation: \hfill 40.6 cm ($\sigma$: 8.2)\\ Maize GDD: \hfill 2,154.4 $^{\circ}$F ($\sigma$: 102.1)\\ Elevation: \hfill 2,222.0 m ($\sigma$: 371.3)\end{tabular} \\[5mm]  
  \hline  
  \noalign{\vskip2mm}  
  CIB & 7,424.1             & 700-1540            & \begin{tabular}[c]{@{}r@{}}Rooms: \hfill 27,234\\ Farms: \hfill 713\\ Centers: \hfill 78\end{tabular}    & \begin{tabular}[r]{@{}r@{}}Precipitation: \hfill 34.3 cm ($\sigma$: 7.6)\\ Maize GDD: \hfill 2,317.3 $^{\circ}$F ($\sigma$: 93.8)\\ Elevation: \hfill 2,145.9 m ($\sigma$: 160.0)\end{tabular} \\[5mm] 
  \hline
  \noalign{\vskip2mm}
\end{tabular}
\end{singlespace}

```

Regional Summaries

:::

@@

Our clustering algorithm draws inspiration from the procedure used by the United States Office of Management and Budget to define Core Based Statistical Areas (CBSA) [@usomb2021]. We start by defining a core area, in this case a community center location. We then associate neighboring residential sites or farmsteads with this center based on their commute times, and we merge core areas into larger agglomerations based on the proportion of their populations that they share. Our method may be described as a guided density-based clustering algorithm in that it does not rely on a random set of points when initialized. Instead, the selection of core areas is guided by archaeological data and regional expertise. We think this is a key argument in its favor.

From an archaeological perspective, one can think of our algorithm as combining insights from @varien1999 and @reese_etal2019, specifically Varien’s idea of a community catchment, which is an isochrone defined by a uniform commute time in all directions from a community center, and Reese’s suggestion that a spatial or geographic community can be identified by grouping individuals in terms of the commute distances between them. Our way of combining these ideas is to associate farm sites with specific community centers in terms of their commute distances and then draw the community catchment around those associated farms. We also share with them and others [@lipe_hegmon1989; @mahoney_etal2000; @peterson_drennan2005] a focus on "geographic" as opposed to "purely social" communities; that is, communities constrained in space and comprised of individuals that can be expected to interact on a regular, even daily, basis.

In broad outline, our proposed community detection algorithm has the following steps:

1. Identify community centers
2. Identify farms within a commute time threshold ($D{\text -}max$)
3. Join communities to their overlapping neighbors ($P$, $D{\text -}join$)
4. Exclude farms closer to another community center (tie-breaking step)
5. Draw smallest concave hull encompassing all farms, centers, and paths

@@

The primary tuning parameters in our algorithm are $P$, the proportion of a community population used to join communities; $D{\text -}join$, the commute time required to calculate P; and $D{\text -}max$, the commute time used to exclude distant farms. These are described in more detail below.

@@

Why persistent interactions? Intuitively, a community center is a place where individuals from different households cross paths with the intention to interact. A simple way to demonstrate that intention is to show that individuals would still be willing to pay some non-negligible cost to visit those centers under various counterfactual changes to local conditions. This serves to distinguish them from merely accidental exchanges or chance encounters. Of course, there can be some path-dependence to this process, with locations of ephemeral or one-off encounters coming to be locations of persistent interaction. The point is that the location will come to be a community center so long as individuals continue to visit that location despite the underlying variance in their circumstances.

Why unrelated individuals? The simple answer is that a community is not a family. As a unit of social organization, it does share something in common with a family, namely, a set of shared interests; but unlike in a family, those shared interests form only a small part of a much larger set of potentially conflicting goals and desires. As a consequence, communities are more susceptible to problems with coordination or collective action. On the other hand, the mere existence of a community center implies some level of coordination - even if that coordination is just about the location of the community center itself! So, community members share less in common than families, but more in common than a purely random sample of the larger population.

We can summarize these ideas by noting that persistent interactions at a location make that location a *center*, and persistent interactions between unrelated individuals at that location make it a *community* center. To make these considerations more useful for defining community boundaries, however, we need to operationalize them, to make them applicable to specific archaeological contexts, and that requires a deeper understanding of the archaeology in those contexts. For the two regions represented by the VEP II dataset, we rely on two different but related sets of criteria to identify community centers. In the Central Mesa Verde area, a site is considered a community center if it includes civic architecture, specifically a great kiva or great house, or eight or more pit structures (small kivas) representing eight or more households. In the Northern Rio Grande, a site counts as a community center if it is classified as a village (a site with 50 or more rooms) or town (a site with 500 or more rooms). In the Cibola region, a community center includes at least 50 rooms or civic architecture. The large number of households in these regions are thought to be too large to be plausibly related by reckoned kinship.

## Join farms within a commute time

**Rule:** For each center, join all farms within commute time, $D{\text -}max$.

Following @reese_etal2019, we define nearness or proximity in terms of commute time rather than geographic distance. This is done for the obvious reason that the primary opportunity cost associated with daily pedestrian movement is the time spent walking, during which other productive activities cannot take place. There is also a much stronger cross-cultural regularity in the length of time individuals devote to commuting than the distance covered, as the latter is a function of the speed of movement, which is influenced by transport technology [@marchetti1994]. As a result, both the frequency and intensity of interaction at community centers should decay with commute time. The longer it takes to get to a community center, the less frequent and less intense the interactions at that center should be [@varien1999; @peterson_drennan2005].

@@

The concept of commute time is a critical variable in many places in our algorithm, but as a first approximation, we incorporate the idea by setting a commute time threshold, $D{\text -}max$. This variable, in effect, defines an isochrone around each community center, what @varien1999 calls a "catchment." All farm sites that fall within that isochrone get associated with its community center. This threshold also serves as a filter, as we drop all farm sites that are farther than that commute time to any center. Doing so helps to minimize potential overestimates of total arable land area for each community. Here, we have defined $D{\text -}max$ as one hour. This is about twice the median daily commute time observed in cross-cultural studies and it includes roughly 95% of the farmsteads in each region (see @fig-commutes).

For details of how we estimate commute times over costly terrains, see “Least-cost path analysis,” below.

@@

**Rule:** For centers $C_1$ and $C_2$ with nearby farm populations $N_1 ≤ N_2$, if $P \cdot N_1$ are within distance $D{\text -}join$ of $C_1$ and $C_2$, then join $C_1$ to $C_2$.

The larger and more dense farming populations become, the harder it gets to tell them apart and, more importantly, the harder it gets to justify keeping them apart, so we need to articulate a rule that specifies when two communities should be merged into one. For CBSAs, that decision hinges on a distinction between central and outlying counties, or counties where some fraction of the population live in an urban core and counties where some fraction commute to and from a central county on a regular basis. Basically, if the central counties of one CBSA qualify as the outlying counties of another CBSA, then the two CBSAs are merged. The intuition here is that when rates of interaction and exchange between CBSAs are roughly equal to rates within a CBSA, the CBSAs should be merged (USOMB 2021).

@@

More precisely, the rule requires that $P \cdot N_1$ be within $D{\text -}join$ of *both* $C_1$ and $C_2$. This promotes joins when $C_1$ and $C_2$ are themselves in close proximity to each other, as the farther apart they are, the smaller the number of sites that overlap within $D{\text -}join$. It also tends to encourage joining two communities when a substantial proportion of the population lives between the two centers. In either case, the result is what @varien1999 sometimes refers to as a "macro-community" or "multi-community cluster," meaning a community organized around multiple community centers.

The rule was partly inspired by the density-based clustering algorithm known as DBSCAN [@ester_etal1996], as we rely on a moving window (an isochrone, in this case) set around each center, and look for some measure of density to guide join-decisions. As noted above, however, our method does not rely on a random initialization of points, but rather targets community centers. It is also hierarchical in the sense that it exploits additional structure in our data, namely the distinction between community centers and farm sites.

>>>>

---
title: A method for defining dispersed community territories
date: today
author:
  - name: Kenneth B. Vernon
    email: Kenneth.Vernon@colorado.edu
    affiliations: 
      - id: colorado
        address: | 
          Center for Collaborative Synthesis in Archaeology,  
          Institute of Behavioral Science,  
          University of Colorado, Boulder,  
          1440 15th Street,  
          Boulder, CO 80302
    attributes:
        corresponding: true 
  - name: Scott G. Ortman
    affiliations:
      - ref: colorado
abstract: |
  The transition from dispersed to aggregated forms of settlement reflects a critical shift in the relative value of social and primary (food) modes of production. However, investigating trade-offs between them requires estimates of the extent of community territories, including their nearby arable land. Here we demonstrate a simple algorithm to do that. Our algorithm is analogous to that used to define core-based statistical areas for the US census, though instead of modern metropolitan areas, we rely on community centers (or areas of known and persistent interaction between unrelated individuals). We provide examples of our algorithm by applying it to archaeological sites in the central Mesa Verde, northern Rio Grande, and Cibola regions. A sensitivity analysis is also conducted to demonstrate how each tuning parameter contributes to the algorithm.
keywords:
  - Aggregation
  - Core-Based Statistical Area
  - Community Detection
  - Complex Systems
  - Least-Cost Path Analysis
  - Population Density
target-journal: Journal of Archaeological Science
bibliography: bibliography.bib
corresponding_author: Kenneth.Vernon@colorado.edu
linenumbers: false
format: 
  simple-pdf: default
  docx: default
execute: 
  echo: false
  error: false
  # warning: false
engine: knitr
---

@@

From a complex systems perspective, the main way these two settlement
morphologies vary is with respect to which costs are being minimized. In the
dispersed pattern, costs associated with primary (food) production by the
household are being minimized, whereas in the aggregated pattern, it is costs
associated with various forms of social production, from government and ritual
to economic exchange and warfare. Presumably, choices regarding which sorts of
costs to minimize are driven by the relative productivity of primary and social
production, which is to say, their relative importance, broadly construed, for
human well-being in a given context.

@@

The datasets we apply our algorithm to come from three regions in the US
Southwest: Central Mesa Verde (CMV), the Northern Rio Grande (NRG), and the
Cibola or Zuni region (CIB) (see @fig-labeled-overview). The CMV and NRG
datasets derive from Phase II of the Village Ecodynamics Project [@ortman2016;
@schwindt_etal2016]. The CIB data derive from data syntheses produced by
Schachner and Peeples [@peeples2018; @peeples_schachner2012; @schachner2012].
All of these data are now included in cyberSW [@mills_etal2020]. The
distribution of farms and centers is shown in @fig-overview. For summaries of
the archaeological and environmental context in each region, see @tbl-regions.

::: {#fig-labeled-overview fig-align="center"}
![](../figures/labeled-overview-map.png){width="5.2in"}

Overview map showing the locations of the three study areas: Central Mesa Verde,
Northern Rio Grande, and Cibola. Prominent towns, administrative units, and
landforms are also labeled to help orient the reader.
:::

The CMV area encompasses the southwestern corner of Colorado, from Mesa Verde
National Park in the southeast to the Dolores River in the north, and the Utah
state line in the west. The elevation ranges from roughly 1,399.8 m to 3,032.8
m, with an average elevation of 1,999.0 m ($\sigma$ = 276.5 m). SKOPE
[@bocinsky_etal2022] estimates suggest annual precipitation in the CMV study
area has averaged around 39.5 cm ($\sigma$ = 9.1 cm) over the last 2,000 years,
and it puts summer growing degree days for maize around 2,390.1 °F ($\sigma$ =
97.9 °F). Across most areas in the region, it is believed that direct
precipitation was sufficiently high to support dry farming in most years
[@bocinsky_kohler2014; @bocinsky_etal2016].

The NRG study area in north central New Mexico includes basically all of the Rio
Grande valley between the Jemez Mountains to the west and the southernmost
mountains of the Sangre De Cristos to the east, with the Rio Grande itself
running from the northeast down to the western edge of the valley at the foot of
the Pajarito Plateau around Bandelier National Monument. The elevation ranges
from roughly 1,587.9 m to 3,836.2 m, with an average elevation of 2,220.0 m
($\sigma$ = 371.3 m). SKOPE estimates the average annual precipitation in the
NRG over the last 2,000 years to be around 40.6 cm ($\sigma$ = 8.2 cm), and it
puts summer growing degree days for maize around 2,154.4 °F ($\sigma$ = 102.1
°F). It is argued that direct rainfall at Bandelier NM made dry farming possible
there, but in the valley farming required floodwater irrigation along
watercourses [@duwe_anschuetz2013; @bocinsky_kohler2014; @bocinsky_etal2016;
@ortman2016a].

The CIB region straddles the southern slope of the Colorado Plateau along the
border between Arizona and New Mexico, from El Morro National Monument in the
east to the centrally located Zuni Pueblo to the confluence of Jaralosa Draw and
the Zuni River in the southwest. Although this is a rugged terrain by any
reasonable standard, it is notably less so than the other study areas. The
elevation ranges from roughly 1,771.3 m to 2,777.1 m, with an average elevation
of 2,145.9 m ($\sigma$ = 160.0 m). SKOPE estimates suggest annual precipitation
in the CIB area over the last 2,000 years has averaged around 34.3 cm ($\sigma$
= 7.6 cm), and it puts summer growing degree days for maize around 2,317.3 °F
($\sigma$ = 93.8 °F). However, no area in the region receives sufficient direct
rainfall to support dry farming, so some form of water management was required
[@kintigh1985; @muenchrath_etal2002].

::: {#tbl-regions}
```{=latex}

\begin{singlespace}
\small
\begin{tabular}{lllll}
    & 
\textbf{Area [km2]} & 
\textbf{Dates [CE]} & 
\textbf{Archaeology} & 
\textbf{Environment} 
  \\ 
  \hline  
  \noalign{\vskip2mm}  
CMV & 
4,566.4 & 
725-1280 & 
\begin{tabular}[c]{@{}r@{}}
  Rooms: \hfill 68,940\\ 
  Farms: \hfill 5,272\\ 
  Centers: \hfill 170
\end{tabular} & 
\begin{tabular}[r]{@{}r@{}}
  Precipitation: \hfill 39.5 cm ($\sigma$: 9.1)\\ 
  Maize GDD: \hfill 2,390.1 $^{\circ}$F ($\sigma$: 97.9)\\ 
  Elevation: \hfill 1,999.0 m ($\sigma$: 276.5)
\end{tabular} 
  \\[5mm]  
  \hline  
  \noalign{\vskip2mm}  
NRG & 
6,958.0 & 
900-1550 & 
\begin{tabular}[c]{@{}r@{}}
  Rooms: \hfill 46,226\\ 
  Farms: \hfill 2,316\\ 
  Centers: \hfill 181
\end{tabular} & 
\begin{tabular}[r]{@{}r@{}}
  Precipitation: \hfill 40.6 cm ($\sigma$: 8.2)\\ 
  Maize GDD: \hfill 2,154.4 $^{\circ}$F ($\sigma$: 102.1)\\ 
  Elevation: \hfill 2,222.0 m ($\sigma$: 371.3)
\end{tabular} 
  \\[5mm]  
  \hline  
  \noalign{\vskip2mm}  
CIB & 
7,424.1 & 
700-1540 & 
\begin{tabular}[c]{@{}r@{}}
  Rooms: \hfill 26,818\\ 
  Farms: \hfill 713\\ 
  Centers: \hfill 77
\end{tabular} & 
\begin{tabular}[r]{@{}r@{}}
  Precipitation: \hfill 34.3 cm ($\sigma$: 7.6)\\ 
  Maize GDD: \hfill 2,317.3 $^{\circ}$F ($\sigma$: 93.8)\\ 
  Elevation: \hfill 2,145.9 m ($\sigma$: 160.0)
\end{tabular} 
  \\[5mm] 
  \hline
  \noalign{\vskip2mm}
\end{tabular}
\end{singlespace}
```
Regional Summaries
:::

@@

Our clustering algorithm draws inspiration from the procedure used by the United
States Office of Management and Budget to define Core Based Statistical Areas
(CBSA) [@usomb2021]. We start by defining a core area, in this case a community
center location. We then merge core areas into larger agglomerations based on
the proportion of their populations that they share. Our method may be described
as a guided density-based clustering algorithm in that it does not rely on a
random set of points when initialized. Instead, the selection of core areas is
guided by archaeological data and regional expertise. We think this is a key
argument in its favor.

From an archaeological perspective, one can think of our algorithm as combining
insights from @varien1999 and @reese_etal2019, specifically Varien's idea of a
community catchment, which is an isochrone defined by a uniform commute time in
all directions from a community center, and Reese's suggestion that a spatial or
geographic community can be identified by grouping individuals in terms of the
commute distances between them. Our way of combining these ideas is to associate
farm sites with specific community centers in terms of their commute distances
and then draw the community catchment around those associated farms. We also
share with them and others [@lipe_hegmon1989; @peterson_drennan2005;
@murdock1949] a focus on "geographic" as opposed to "purely social" communities;
that is, communities constrained in space and comprised of individuals that can
be expected to interact on a regular, even daily, basis.

An important difference for our algorithm is that it incorporates information
about the total mass or population size around each center, so it is at least
conceptually similar to the classic Xtent model proposed by [@renfrew_level1979;
see also @ducke_kroefges2008], as well as the more widely known gravity model
[@isard1954]. In broad outline, it has the following steps:

1.  Identify community centers
2.  Join farms to their nearest community center
3.  Exclude farms beyond a commute time threshold from all centers
    ($D{\text -}max$)
4.  Join centers based on their overlapping populations ($P$, $D{\text -}join$)
5.  Draw smallest concave hull encompassing all farms, centers, and paths

@@

The primary tuning parameters in our algorithm are $P$, the proportion of a
community population used to join communities; $D{\text -}join$, the commute
time required to calculate P; and $D{\text -}max$, the commute time used to
exclude distant farms. While values for these variables were carefully chosen
based on theory and empirical research, we recognize that they are, at the end
of the day, arbitrary selections, so we include some limited sensitivity
analysis to show how changes to these parameters affect the algorithm (see
@fig-sensitivity). This involves changing the value of a focal parameter while
holding the other parameters at their default values, which we describe below.

::: {#fig-overview fig-align="center"}
![](../figures/overview-maps.png){width="6.5in"}

Overview map showing the locations of farmsteads (black dots) and community
centers (red triangles) in Central Mesa Verde, Northern Rio Grande, and Cibola.
:::

@@

Why persistent interactions? Intuitively, a community center is a place where
individuals from different households cross paths with the intention to interact
[@glowacki_ortman2012; @varien1999; @berrey_etal2021]. A simple way to
demonstrate that intention is to show that individuals would still be willing to
pay some non-negligible cost to visit those centers under various counterfactual
changes to local conditions. This serves to distinguish them from merely
accidental exchanges or chance encounters. Of course, there can be some
path-dependence to this process, with locations of ephemeral or one-off
encounters coming to be locations of persistent interaction. The point is that
the location will come to be a community center so long as individuals continue
to visit that location despite the underlying variability in their
circumstances.

Why unrelated individuals? The simple answer is that a community is not a
family. As a unit of social organization, it does share something in common with
a family, namely, a set of shared interests; but unlike in a family, those
shared interests form only a small part of a much larger set of potentially
conflicting goals and desires. As a consequence, communities are more
susceptible to problems of coordination or collective action [@smith2003;
@smith2010]. On the other hand, the mere existence of a community center implies
some level of coordination - even if that coordination is just about the
location of the community center itself! So, community members share less in
common than families, but more in common than a purely random sample of the
larger population.

We can summarize these ideas by noting that persistent interactions at a
location make that location a *center*, and persistent interactions between
unrelated individuals at that location make it a *community* center
[@varien_potter2008; @gilpin2003; @stone2016; @peterson_drennan2005]. To make
these considerations more useful for defining community boundaries, however, we
need to operationalize them, to make them applicable to specific archaeological
contexts, and that requires a deeper understanding of the archaeology in those
contexts. For given differences in population density and settlement patterning,
the frequency and intensity of interaction between individuals at centers is
likely to vary by region, so it is reasonable to expect that community centers
themselves (at least in so far as they play the role of centers) should vary in
their size and composition across regions, too.

For the two regions represented by the VEP II dataset, we rely on two different
but related sets of criteria to identify community centers. In the Central Mesa
Verde area, a site is considered a community center if it includes civic
architecture, specifically a great kiva or great house, or eight or more pit
structures (small kivas) representing eight or more households. In the Northern
Rio Grande, a site counts as a community center if it is classified as a village
(a site with 50 or more rooms) or town (a site with 500 or more rooms). In the
Cibola region, a community center includes at least 50 rooms or civic
architecture. The large number of households in these regions are thought to be
too large to be plausibly related by reckoned kinship.

## Join farms to their nearest community center

**Rule:** For each farm, assign it to the community of its nearest community
center.

Following @reese_etal2019, we define nearness or proximity in terms of commute
time rather than linear geographic distance. This is done for the obvious reason
that the primary opportunity cost associated with daily pedestrian movement is
the time spent walking, during which other productive activities cannot take
place. There is also a much stronger cross-cultural regularity in the length of
time individuals devote to commuting than the distance covered, as the latter is
a function of the speed of movement, which is influenced by transport technology
[@marchetti1994]. As a result, both the frequency and intensity of interaction
at community centers should decay with commute time. The longer it takes to get
to a community center, the less frequent and less intense the interactions at
that center should be [@varien1999; @peterson_drennan2005].

@@

Commute time is a critical variable in many places in our algorithm, but as a
first approximation, we incorporate the idea by simply joining farms to the
community of their nearest center, meaning the center with the shortest commute
time from the farm. For details of how we estimate commute times over costly
terrains, see "Least-cost path analysis," below.

## Exclude distant farms

**Rule:** Exclude all farms farther than commute time, $D{\text -}max$, from
their nearest center.

The commute time threshold, $D{\text -}max$, defines an isochrone around each
community center. All farm sites that fall outside that isochrone are considered
outliers and dropped from the analysis. Doing this helps to minimize potential
overestimates of total arable land area for each community. This is owing to the
fact that distant farms on average add more area to the community polygon than
nearer farms. Here, we have defined $D{\text -}max$ as one hour. This is about
twice the median daily commute time observed in cross-cultural studies and it
includes roughly 95% of the farmsteads in each region (see @fig-commutes).

@@

**Rule:** For any two centers $c$ and $d$ with populations $N_{c} < N_{d}$, if
$P \cdot N_{c}$ is within distance $D{\text -}join$ of $d$, then $c$ is part of
the same community as $d$.

The larger and more dense farming populations become, the harder it gets to tell
them apart and, more importantly, the harder it gets to justify keeping them
apart, so we need to articulate a rule that specifies when two communities
should be merged into one. For CBSAs, that decision hinges on a distinction
between central and outlying counties, or counties where some fraction of the
population live in an urban core and counties where some fraction commute to and
from a central county on a regular basis. Basically, if the central counties of
one CBSA qualify as the outlying counties of another CBSA, then the two CBSAs
are merged. The intuition here is that when rates of interaction and exchange
between CBSAs are roughly equal to rates within a CBSA, the CBSAs should be
merged [@usomb2021].

@@

Here we operationalize the population size $N$ of center $c$ using the
well-established rule of thumb in US Southwest archaeology that each surface
room in a settlement represents a single person, leading to a one to one
conversion from rooms to residents [@duwe_etal2016; @kintigh1985; @lekson1989;
@lipe1989; @ortman2016]. In this case, we use the catchment room count, meaning
the number of rooms within $D{\text -}join$ of $c$, defined as:

$$N_{c} = \sum_{i=1}^{S} R_{i} \cdot I(t_{ic} \leq D{\text -}join)$$

\noindent for all sites $S$ (including both farms and centers), with $R_{i}$
being the room count at site $i$, $t_{ic}$ the travel time from $i$ to $c$,
$D{\text -}join$ the threshold travel time to a center, and $I$ the indicator
function that is 1 if $t \leq D{\text -}join$ and 0 otherwise. Note that $S$
includes $c$ and that $t_{cc} = 0$ (the travel time from a center to the same
center is precisely 0), so $R_{c}$ is always included in $N_{c}$.

The catchment room count within $D{\text -}join$ of two centers $c$ and $d$ is
then

$$N_{cd} = \sum_{i=1}^{S} R_{i} \cdot I(t_{ic} \leq D{\text -}join) \cdot I(t_{id} \leq D{\text -}join)$$

\noindent and

$$P_{cd} = N_{cd}/N_{c}$$
